---
title: "dodgr, Distances on Directed Graphs in R"
output:
  rmarkdown::html_vignette:
    self_contained: no

  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r opts, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = TRUE,
  message = TRUE,
  width = 120,
  comment = "#>",
  fig.retina = 2,
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/mpadge/dodgr.svg)](https://travis-ci.org/mpadge/dodgr) 
[![codecov](https://codecov.io/gh/mpadge/dodgr/branch/master/graph/badge.svg)](https://codecov.io/gh/mpadge/dodgr)
[![Project Status: Concept - Minimal or no implementation has been done yet.](http://www.repostatus.org/badges/0.1.0/concept.svg)](http://www.repostatus.org/#concept)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/dodgr)](http://cran.r-project.org/web/packages/dodgr) 

# dodgr: Distances on Directed Graphs in R

R package for calculating pairwise distances on dual-weighted directed graphs
using Priority Queue Shortest Paths. Dual-weighted directed graphs are directed
graphs with two sets of weights so that `weight1(A->B) != weight1(B->A)`---the
directed property---and `weight2(A->B) != weight1(A->B)`---the dual property.
`dodgr` calculates shortest paths according to one weight, while distances along
paths are calculating using the other weight. A canonical example of a
dual-weighted directed graph is a street network to be used for routing. Routes
are usually calculated by weighting different kinds of streets or ways according
to a particular mode of transport, while the desired output is a direct distance
which has to be calculated using a different set of weights.

The shortest path algorithm relies on priority queue code adapted from original
code by Shane Saunders available here
[http://www.cosc.canterbury.ac.nz/tad.takaoka/alg/spalgs/spalgs.html](http://www.cosc.canterbury.ac.nz/tad.takaoka/alg/spalgs/spalgs.html)

## Installation

You can install `dodgr` from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mpadge/dodgr")
```
```{r load-pqsqpr, echo = FALSE, message = FALSE}
devtools::load_all (".", export_all = FALSE)
```

## Simple Usage

The primary function,
```{r, eval = FALSE}
d <- dodgr_dists (graph = graph, from = from)
```
produces a square matrix of distances between all points listed in `from` and
routed along the dual-weighted directed network given in `graph`. An even
simpler usage allows calculation of pair-wise distances between a set of
geographical coordinates (here, for a sizey chunk of New York City):
```{r ny-example}
xlim <- c (-74.12931, -73.99214)
ylim <- c (40.70347, 40.75354)
npts <- 1000
pts <- data.frame (x = xlim [1] + runif (npts) * diff (xlim),
                   y = ylim [1] + runif (npts) * diff (ylim))
st <- Sys.time ()
d <- dodgr_dists (from = pts)
Sys.time () - st
range (d, na.rm = TRUE)
```
This will automatically download the street network (using
[`osmdata`](https://cran.r-project.org/package=osmdata)), and even then
calculating distances between 1,000 points can be done in around 10 seconds.

## Detailed Usage

A graph or network in `dodgr` is represented as a flat table
(`data.frame`, `tibble`, `data.table`, whatever) of minimally four columns:
`from`, `to`, `weight`, and `distance`. The first two can be of arbitrary form
(`numeric` or `character`); `weight` is used to evaluate the shortest paths, and
the desired distances are evaluated by summing the values of `distance` along
those paths. For a street network example, `weight` will generally be the actual
distance multiplied by a priority weighting for a given mode of transport and
type of way, while `distance` will be the pysical distance.

The primary function of `dodgr` is `dodgr_dists()`, which accepts the three main
arguments of

1. `graph` - the flat table described above;
2. `from` - a list of origin points for which distances are to be calcualted; and
3. `to` - a list of equivalent destination points. If missing, pairwise
   distances are calculated between all points in `from`.

For spatial graphs in which `graph` has additional longitude and latitude
coordinates, `from` and `to` may also be given as lists of coordinates;
otherwise they must match onto names or values given in `graph$from` and
`graph$to`. Coordinates in `from` and `to` that do not exactly match on to
values given in `graph` will be mapped on to the nearest `graph` nodes.
