---
title: "pqspr, Priority Queue Shortest Paths in R"
output:
  rmarkdown::html_vignette:
    self_contained: no

  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r opts, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = TRUE,
  message = TRUE,
  width = 120,
  comment = "#>",
  fig.retina = 2,
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/mpadge/pqspr.svg)](https://travis-ci.org/mpadge/pqspr) 
[![Project Status: Concept - Minimal or no implementation has been done yet.](http://www.repostatus.org/badges/0.1.0/concept.svg)](http://www.repostatus.org/#concept)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/pqspr)](http://cran.r-project.org/web/packages/pqspr) 
![downloads](http://cranlogs.r-pkg.org/badges/grand-total/pqspr)

# pqspr

Priority Queue Shortest Paths in R. Simply because they are blindingly faster
than anything else, yet do not seem to be implemented in any current package.
Note that most of the source code which does the work here is only lightly
adapted from original code of Shane Saunders obtained from
[http://www.cosc.canterbury.ac.nz/tad.takaoka/alg/spalgs/spalgs.html](http://www.cosc.canterbury.ac.nz/tad.takaoka/alg/spalgs/spalgs.html)

## Installation

You can install `pqspr` from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mpadge/pqspr")
```
```{r load-pqsqpr, echo = FALSE, message = FALSE}
devtools::load_all (".", export_all = FALSE)
```

## Timing Comparison

Create a street network using `osmdata` and some code currently in the github
repo [`osmprob`](https://github.com/osm-router/osmprob) which is installed
first.
```{r, eval = FALSE}
devtools::install_github ("osm-router/osmprob")
```
```{r load-osmprob, echo = FALSE, message = FALSE}
devtools::load_all ("../osmprob", export_all = FALSE)
```
Use `osmprob` to get a street network graph:
```{r download-graph}
start_pt <- c (-74.00150, 40.74178)
end_pt <- c (-74.07889, 40.71113)
graph <- download_graph (start_pt, end_pt)
```
The result has a `$original` graph containing all vertices and a `$compact`
graph with redundant vertices removed. We'll perform routing on the latter which
has this many edges:
```{r}
nrow (graph$compact)
```
Both of these graphs are simple data frames detailing all edges.  Some light
wrangling is now necessary to prepare both the `igraph` and the simple structure
submitted to the `pqspr` routine:
```{r igraph-setup}
edges <- cbind (graph$compact$from_id, graph$compact$to_id)
nodes <- unique (as.vector (edges)) # used below in test comparison
edges <- as.vector (t (edges))
igr <- igraph::make_directed_graph (edges)
igraph::E (igr)$weight <- graph$compact$d_weighted

graph <- graph$compact
indx <- which (names (graph) %in% c ("from_id", "to_id", "d", "d_weighted"))
graph <- graph [, indx]
graph$from_id <- paste0 (graph$from_id)
graph$to_id <- paste0 (graph$to_id)
```
Then the final timing comparison between `igraph::distances()`, which returns a
matrix of distances between all vertices, and the equivalent and only
function of this package, `test()`:
```{r, benchmark}
rbenchmark::benchmark (test (graph),
                   igraph::distances (igr, v = nodes, to = nodes, mode = "out"),
                       replications = 10)
```
And these priority queue routines are over ten times faster than the `igraph`
equivalent.
