# script to generate `data/hampi.rda` 

```{r load, echo = FALSE}
library (magrittr)
library (osmdata)
```
```{r}
hampi <- opq ("hampi india") %>%
    add_osm_feature (key = "highway") %>%
    osmdata_sf () %>%
    osm_poly2line () %>%
    extract2 ("osm_lines")
```
Then need to get rid of columns, especially the ones with Kannada names which
are non-UTF8
```{r}
nms <- c ("osm_id", "bicycle", "covered", "foot", "highway", "incline",
          "motorcar", "motorcycle", "motor_vehicle", "oneway", "surface",
          "tracktype", "tunnel", "width", "geometry")
hampi <- hampi [, match (nms, names (hampi))]
```
```{r}
devtools::use_data (hampi, overwrite = TRUE, compress = 'xz')
```

# generate `data/weight_profiles`

additional values from OSRM profiles at
https://github.com/Project-OSRM/osrm-backend/blob/master/profiles/

```{r}
library (magrittr)
theurl <- "https://www.routino.org/xml/routino-profiles.xml"
dat <- xml2::read_html (theurl) %>%
    rvest::html_nodes("profile") %>%
    xml2::as_list ()
```
```{r}
weighting_profiles <- lapply (dat, function (i) {
                 di <- i$preferences
                 res <- lapply (di, function (j) 
                         c (attr (i, "name"),
                            attr (j, "highway"),
                            attr (j, "percent")))
                 do.call (rbind, res)
          })
# Then add living_street, bridleway, and footway to all profiles
# https://wiki.openstreetmap.org/wiki/Tag:highway%3Dliving_street
nms <- sapply (weighting_profiles, function (i) i [1, 1])

#               living_street   footway     bridleway
# "foot"        95              100         100
# "horse"       80              100         100
# "wheelchair"  95              100         50
# "bicycle"     95              90          70
# "moped"       60              0           0
# "motorcycle"  50              0           0
# "motorcar"    40              0           0
# "goods"       30              0           0
# "hgv"         30              0           0
# "psv"         30              0           0

wt_ls <- c (95, 80, 95, 95, 60, 50, 40, 30, 30, 30) # living street
wt_br <- c (100, 100, 50, 70, 0, 0, 0, 0, 0, 0) # bridleway
wt_fw <- c (100, 100, 100, 90, 0, 0, 0, 0, 0, 0) # bridleway
names (wt_ls) <- names (wt_br) <- names (wt_fw) <- nms
newrows <- data.frame ("living_street" = wt_ls,
                       "bridleway" = wt_br,
                       "footway" = wt_fw)
weighting_profiles <- lapply (weighting_profiles, function (i) {
            newdat <- t (newrows [rownames (newrows) == i [1, 1], ])
            newdat <- cbind (i [seq (ncol (newrows)), 1],
                             rownames (newdat),
                             newdat [, 1])
            rbind (i, newdat)
                       })

# plus all of the "link" types defaulting to same as parent entities
links <- c ("motorway", "trunk", "primary", "secondary", "tertiary")
nms <- paste0 (links, "_link")
weighting_profiles <- lapply (weighting_profiles, function (i)
                              {
                                  index <- match (links, i [, 2])
                                  newrows <- cbind (i [index, 1],
                                                    nms,
                                                    i [index, 3])
                                  res <- rbind (i, newrows)
                                  rownames (res) <- rep ("preference",
                                                         nrow (res))
                                  return (res)
                              })

weighting_profiles <- do.call (rbind, weighting_profiles)
weighting_profiles <- data.frame (name = weighting_profiles [, 1],
                           way = weighting_profiles [, 2],
                           value = as.numeric (weighting_profiles [, 3]),
                           stringsAsFactors = FALSE)
```

### maxspeed column + traffic light and turn penalties

Values adopted from the OSRM profiles, all in km / hr

maxspeeds:
```{r}
speeds <- rbind (c ("bicycle", 15),
                 c ("foot", 5),
                 c ("wheelchair", 5),
                 c ("horse", 6.4),
                 c ("moped", 26),
                 c ("motorcar", 36),
                 c ("motorcycle", 36),
                 c ("goods", 36),
                 c ("hgv", 36),
                 c ("psv", 36))
speeds <- data.frame ("name" = speeds [, 1],
                      "speed" = as.numeric (speeds [, 2]))
```

surface speeds as proportions of full speed:
```{r}
s <- rbind (c ("bicycle", "cobblestone", "flattened", 0.75),
            c ("bicycle", "surface", "paving_stones", 0.75),
            c ("bicycle", "surface", "compacted", 0.75),
            c ("bicycle", "surface", "cobblestone", 0.4),
            c ("bicycle", "surface", "unpaved", 0.4),
            c ("bicycle", "surface", "fine_gravel", 0.4),
            c ("bicycle", "surface", "gravel", 0.4),
            c ("bicycle", "surface", "pebblestone", 0.4),
            c ("bicycle", "surface", "ground", 0.4),
            c ("bicycle", "surface", "dirt", 0.4),
            c ("bicycle", "surface", "earth", 0.4),
            c ("bicycle", "surface", "grass", 0.4),
            c ("bicycle", "surface", "mud", 0.2),
            c ("bicycle", "surface", "sand", 0.2),
            c ("bicycle", "surface", "sett", 0.75),

            c ("foot", "surface", "fine_gravel", 0.75),
            c ("foot", "surface", "gravel", 0.75),
            c ("foot", "surface", "pebblestone", 0.75),
            c ("foot", "surface", "mud", 0.5),
            c ("foot", "surface", "sand", 0.5))
surface_speeds <- data.frame ("name" = s [, 1],
                              "key" = s [, 2],
                              "value" = s [, 3],
                              "speed" = s [, 4],
                              stringsAsFactors = FALSE)
                              
```

speeds on other ways (not implemented)
```{r, eval = FALSE}
bike_walking <- 4
bike_walking_steps <- 2
```

traffic light and turn penalties. These are calculated in OSRM based on turn
angles. The `dodgr` algorithm uses turn ordering across traffic, and so only
needs to implement a fixed traffic light waiting time cost. The equivalent
values can be obtained from the OSRM profiles with a turn angle of 0 as:
```{r traffic-light-penalties, eval = FALSE}
traffic_light_penalty <- 2
turn_duration = traffic_light_penalty
turn_bias <- 1.075 # car
turn_duration <- turn_duration + 6.5 / turn_bias # = 8

# for both bike and car:
turn_duration <- traffic_light_penalty # = 2
```

```{r}
nms <- c ("foot", "horse", "wheelchair", "bicycle", "moped", "motorcycle",
          "motorcar", "goods", "hgv", "psv")
traffic_lights <- rep (8, length (nms))
traffic_lights [nms %in% c ("foot", "wheelchair", "bicycle")] <- 2
turn_penalty <- rep (7.5, length (nms))
turn_penalty [nms %in% c ("foot", "horse", "wheelchar")] <- 0
turn_penalty [nms == "bicycle"] <- 6
penalties <- data.frame ("name" = nms,
                         "traffic_lights" = traffic_lights,
                         "turn" = turn_penalty,
                         stringsAsFactors = FALSE)
```

Note that the following link describes waiting times at traffic lights with an
average value of 40s:
http://dx.doi.org/10.1016/j.trb.2013.02.002
see also National Assoc. City Transportation Officials guidelines:
https://nacto.org/publication/urban-street-design-guide/intersection-design-elements/traffic-signals/signal-cycle-lengths/
recommended "Short cycle length" of 60-90 seconds

### save the data:

```{r}
# change percentage value to 0-1
weighting_profiles$value <- weighting_profiles$value / 100
weighting_profiles <- list ("weighting_profiles" = weighting_profiles,
                            "speeds" = speeds,
                            "surface_speeds" = surface_speeds,
                            "penalties" = penalties)
usethis::use_data (weighting_profiles, overwrite = TRUE, compress = 'xz')
```
