---
title: "dodgr benchmark"
author: "Andreas Petutschnig"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
        number_sections: true
        theme: flatly
vignette: >
  %\VignetteIndexEntry{benchmark}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
devtools::load_all (".", export_all = FALSE)
```

# Introduction

This vignette contains runtime comparisons for a number of routing
operations between random vertices using distance calculation functions from the
`dodgr` and [`igraph`](https://cran.r-project.org/package=igraph) packages. The
experiments are run using the `compare_heaps()` function, which successively
calls functions `igraph::distances()` and then `dodgr_dists()` with all
implemented options for the `heap` parameter and measures their respecive
runtimes.

The graph data on which the tests are performed are taken from the
[`igraphdata`](https://cran.r-project.org/package=igraphdata) package.

# Get Graph Data

To be able to run `compare_heaps()`, we need to create `data.frame` objects from
the `igraphdata` data. As we need at least columns for each graph edges' origin,
destination and weight, not all `igraphdata` are fit for our purposes, because
they lack a weight column. In these cases, all their edge weights are assumed to
be 1 so they can be brought into a valid format.

The following code compiles all `igraph` objects into a list of usable
`data.frame` objects.

```{r}
datasets <- data (package = "igraphdata")
datasets <- datasets$results [, "Item"]

all_graphs <- list ()
gr_el <- 1

for (i in seq_along (datasets))
{
    dat <- datasets [i]
    data (package = "igraphdata", list = dat)
    dat_graph <- get (dat)
    if (class (dat_graph) == "igraph")
        dat_graph <- list (dat_graph)
    for (j in seq_along (dat_graph))
    {
        graph <- dat_graph [[j]]
        graph <- igraph::get.data.frame (graph)
        from_id <- graph$from
        to_id <- graph$to
        if ("weight" %in% names (graph))
        {
            d <- graph$weight
            graph_out <- data.frame (from_id, to_id, d)
            all_graphs [[gr_el]] <- graph_out
            gr_el <- gr_el + 1
        } else
        {
            graph [c ("from", "to")] <- NULL
            edge_cols <- which (sapply (graph, class) == "numeric")
            len <- length (edge_cols)
            if (len > 0)
            {
                for (k in seq_len (len))
                {
                    d <- graph [, k]
                    graph_out <- data.frame (from_id, to_id, d)
                    all_graphs [[gr_el]] <- graph_out
                    gr_el <- gr_el + 1
                }
            } else
            {
                d <- rep (1, length (from_id))
                graph_out <- data.frame (from_id, to_id, d)
                all_graphs [[gr_el]] <- graph_out
                gr_el <- gr_el + 1
            }
        }
    }
}
```

Now, `compare_heaps()` can be run for all graphs. First, show the number of
vertices of all graphs in the list.

```{r}
num_vertices <- sapply (all_graphs, nrow)
num_vertices
```

```{r fig.width = 12, fig.height = 8}
# benchmarks <- lapply (all_graphs, compare_heaps)
benchmarks <- readRDS ("benchmark_data.rds")
elapsed <- unlist (purrr::map (benchmarks, "elapsed"))
tests <- unlist (purrr::map (benchmarks, "test"))
tests <- as.character (tests)
tests <- sub ("d <- dodgr_dists\\(graph, from = from_id, to = to_id, heap =",
     "Contracted", tests)
p <- "d <- dodgr_dists\\(graph_contracted, from = from_id, to = to_id, heap = "
tests <- sub (p, "", tests)
tests [grepl ("igraph", tests)] <- "igraph"
tests <- gsub ("\"", "", tests)
tests <- gsub ("\\)", "", tests)
tests <- as.factor (tests)
n_tests <- dim (benchmarks [[1]]) [1]
num_vertices <- rep (num_vertices, each = n_tests)
bm <- data.frame (num_vertices, Test = tests, elapsed)

library (ggplot2)
library (scales)

ggplot (bm) +
    geom_line (aes (x = num_vertices, y = elapsed, colour = Test)) +
    geom_point (aes (x = num_vertices, y = elapsed, colour = Test)) +
    scale_x_log10 (breaks = trans_breaks ("log10", function (x) 10^x),
                   labels = trans_format ("log10", math_format (10^.x))) +
    scale_y_log10 (breaks = trans_breaks ("log10", function (x) 10^x),
                   labels = trans_format ("log10", math_format (10^.x))) +
    labs (title = "dodgr benchmark results",
          x = "Number of vertices (log scale)", y = "Runtime [s] (log scale)") +
    theme_bw (base_family = "TeX Gyre Bonum") +
    theme (plot.title = element_text(hjust = 0.5),
                    legend.box.background = element_rect ())
```
