---
title: "dodgr"
author: "Mark Padgham"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
        number_sections: true
        theme: flatly
header-includes: 
    - \usepackage{tikz}
    - \usetikzlibrary{arrows}
vignette: >
  %\VignetteIndexEntry{dodgr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
devtools::load_all (".", export_all = FALSE)
```

# Background

`dodgr` is an **R** package for calculating **D**istances **O**n **D**irected
**G**raphs. It does so very efficiently, and is able to process much larger
graphs than most other comparable **R** packages. Skip straight to the Intro if
you know what directed graphs are (but maybe make a brief stop-in to
Dual-Weighted Directed Graphs below.) Directed graphs are ones in which the
"distance" (or some equivalent measure) from A to B is not necessarily equal to
that from B to A. In Fig. 1, for example, the weights between the graph vertices
(A, B, C, and D) differ depending on the direction of travel, and it is only
possible to traverse the entire graph in an anti-clockwise direction.

```{r, fig.align="center", echo=FALSE, engine='tikz',fig.ext='svg', code=readLines("fig1.tex"), fig.cap="Figure 1: A weighted, directed graph"}
```

Graphs in `dodgr` are represented by simple flat `data.frame` objects, so the
graph of Fig. 1, presuming the edge weights to take values of 1, 2, and 3, would
be,
```{r, echo = FALSE}
graph <- data.frame (from = c ("A", "B", "B", "B", "C", "C", "D", "D"),
                     to = c ("B", "A", "C", "D", "B", "D", "C", "A"),
                     d = c (1, 2, 1, 3, 2, 1, 2, 1))
graph
```
The primary function of `dodgr` is `dodgr_dists`, which calculates pair-wise
shortest distances between all vertices of a graph.
```{r}
dodgr_dists (graph)
```


## Dual-Weighted Directed Graphs

Shortest-path distances on weighted graphs can be calculated using a number of
other **R** packages, such as
[`igraph`](https://cran.r-project.org/package=igraph) or
[`e1071`](https://cran.r-project.org/package=e1071). `dodgr` comes into its own
through its ability to trace paths through *dual-weighted* directed graphs,
illustrated in Fig. 2.

```{r, fig.align="center", echo=FALSE, engine='tikz',fig.ext='svg', code=readLines("fig2.tex"), fig.cap="Figure 2: A **dual**-weighted, directed graph in which the weights (represented by thicknesses) of the grey arrows differ from those of the black arrows."}
```
Dual-weighted directed graphs are common in many areas, a foremost example
being routing through street networks. Routes through street networks depends on
mode of transport: the route a pedestrian might take will generally differ
markedly from the route the same person might take if behind the wheel of an
automobile. Routing through street networks thus generally requires each edge to
be specified with two weights or distances: one quantifying the physical
distance, and a second weighted version reflecting the mode of transport (or
some other preferential weighting).

`dodgr` calculates shortest paths using one set of weights (called "weights" or
anything else starting with "w"), but returns the actual lengths of them using a
second set of weights (called "distances", or anything else starting with "d").
If no weights are specified, distances alone are used both for routing and final
distance calculations.  Consider that the weights and distances of Fig. 2 are
the black and grey lines, respectively, with the latter all equal to one. In
this case, the graph and associated shortest distances are,
```{r, echo = FALSE}
graph <- data.frame (from = c ("A", "B", "B", "B", "C", "C", "D", "D"),
                     to = c ("B", "A", "C", "D", "B", "D", "C", "A"),
                     w = c (1, 2, 1, 3, 2, 1, 2, 1),
                     d = c (1, 1, 1, 1, 1, 1, 1, 1))
graph
dodgr_dists (graph)
```
Note that even though the shortest "distance" from A to D is actually
A$\to$B$\to$D with a distance of only 2, that path has a weighted distance of
1 + 3 = 4. The shortest *weighted* path is A$\to$B$\to$C$\to$D, with a distance
both weighted and unweighted of 1 + 1 + 1 = 3. Thus `d(A,D) = 3` and not 2.

# Introduction to `dodgr` 

Although the package has been intentionally developed to be adaptable to any
kinds of networks, most of the applications illustrated here concern street
networks, and also illustrate several helper functions the package offers for
working with street networks. The basic `graph` object of `dodgr` is
nevertheless arbitrary, and need only minimally contain three or four columns as
demonstrated in the simple examples at the outset.

The package may be used to calculate a matrix of distances between a given set
of geographic coordinates. We can start by simply generating some random
coordinates, in this case within the bounding box defining the city of York in
the U.K.
```{r get-york-data, eval = FALSE}
bb <- osmdata::getbb ("york uk")
npts <- 1000
xy <- apply (bb, 1, function (i) min (i) + runif (npts) * diff (i))
bb; head (xy)
```
```{r, echo = FALSE}
bb <- rbind (c (-1.241536, -0.9215361), c (53.799056, 54.1190555))
rownames (bb) <- c ("x", "y")
colnames (bb) <- c ("min", "max")
bb
x <- c (-1.1713502, -1.2216108, -1.0457199, -0.9384666, -0.9445541, -1.1207099)
y <- c (53.89409, 54.01065, 53.83613, 53.93545, 53.89436, 54.01262)
cbind (x, y)
```
Those points can simply be passed to `dodgr_dists()`:
```{r dodgr-dists-in-york, eval = FALSE}
system.time(
            d <- dodgr_dists (from = xy, wt_profile = "foot", quiet = FALSE)
            )
```
```{r, echo = FALSE}
message (paste0 ("No graph submitted to dodgr_dists; ",
                 "downloading street network ... done\n",
                 "Converting network to dodgr graph ... done\n",
                 "Calculating shortest paths ... done"))
```
```{r, echo = FALSE}
st <- system.time (Sys.sleep (0.1))
st [1] <- 26.620
st [2] <- 0.132
st [3] <- 28.567
st
```
```{r, eval = FALSE}
dim (d); range (d, na.rm = TRUE)
```
```{r, echo = FALSE}
c (1000, 1000)
c (0.00000, 46.60609)
```
The result is a matrix of 1000-by-1000 distances of up to 47km long, measured
along routes weighted for optimal pedestrian travel. In this case, the single
call to `dodgr_dists()` automatically downloaded the entire street network of York
and calculated one million shortest-path distances, all in under 30 seconds.

# Graphs and Street Networks

Although the above code is short and fast, most users will probably want more
control over their graphs and routing possibilities. Each of the steps indicated
above (through the `quiet = FALSE` option) can be implemented separately. To
begin, let's download a street network:
```{r, york-streetnet, eval = FALSE}
streetnet <- dodgr_streetnet ("york uk")
class (streetnet)
```
```{r, echo = FALSE}
c ("sf", "data.frame")
```
```{r, eval = FALSE}
class (streetnet$geometry)
```
```{r, echo = FALSE}
c ("sfc_LINESTRING", "sfc")
```
```{r, eval = FALSE}
dim (streetnet)
```
```{r, echo = FALSE}
c (18284, 181)
```
The `streetnet` is an [`sf`](https://cran.r-project.org/package=sf) (Simple
Features) object containing 18,284 `LINESTRING` geometries. In other words, it's
got an `sf` representation of 18,284 street segments.  That `sf` representation
can then be converted into a flat `data.frame` object by
```{r, eval = FALSE}
graph <- weight_streetnet (streetnet, wt_profile = "bicycle")
dim (graph)
```
```{r, echo = FALSE}
c (187300, 11)
```
```{r, eval = FALSE}
head (graph)
```
```{r, echo = FALSE}
from_id <- c (13060397, 13060402, 13060402, 13060405, 13060405, 13060413)
from_id <- as.character (from_id)
from_lon <- c ( -1.001714, -1.001607, -1.001607, -1.001372, -1.001372,
               -1.001113)
from_lat <- c ( 53.95737, 53.95746, 53.95746, 53.95763, 53.95763, 53.95776)
to_id <- c (13060402, 13060397, 13060405, 13060402, 13060413, 13060405)
to_id <- as.character (to_id)
to_lon <- c (-1.001607, -1.001714, -1.001372, -1.001607, -1.001113, -1.001372)
to_lat <- c (53.95746, 53.95737, 53.95763, 53.95746, 53.95776, 53.95763)
d <- c (0.006909587, 0.006909587, 0.013732955, 0.013732955, 0.013148026,
        0.013148026)
d_weighted <- c (0.007677319, 0.007677319, 0.015258839, 0.015258839,
                 0.014608918, 0.014608918)
highway <- rep ("unclassified", length (from_id))
component <- rep (1, length (from_id))
data.frame (from_id = as.character (from_id),
                     from_lon = from_lon, from_lat = from_lat,
                     to_id = as.character (to_id),
                     to_lon = to_lon, to_lat = to_lat,
                     d = d, d_weighted = d_weighted, highway = highway,
                     component = component, stringsAsFactors = FALSE)
```
Note that the actual graph contains around 10 times as many edges as there are
streets, indicating that each street contains on average around 11 vertices
or nodes, between which there are 10 edges. These points or vertices can be
extracted with,
```{r, eval = FALSE}
vt <- dodgr_vertices (graph)
head(vt)
```
```{r, echo = FALSE}
id <- c (13060397, 13060402, 13060405, 13060413, 13060417, 13060648)
x <- c (-1.0017135, -1.0016069, -1.0013725, -1.0011133, -1.0005339, -0.9763021)
y <- c (53.95737, 53.95746, 53.95763, 53.95776, 53.95808, 53.96698)
n <- 0:5
vt <- data.frame (id = as.character (id), x = x, y = y, n = n)
rownames (vt) <- c (1, 2, 4, 6, 8, 9)
vt
```
```{r, eval = FALSE}
dim (vt)
```
```{r, echo = FALSE}
c (90015, 4)
```
From which we see that the OpenStreetMap representation of the streets of York
has 18,284 line segments with 90,015 unique points and 187,300 edges between
those points. The number of edges per vertex in the entire network is thus,
```{r, eval = FALSE}
nrow (graph) / nrow (vt)
```
```{r, echo = FALSE}
187300 / 90015
```
A simple straight line has two edges between all intermediate nodes, and this
thus indicates that the network in it's entirety is quite simple. Is is
nevertheless also very large, as will be demonstrated below, extending for
considerable distances beyond the relatively complex inner city, and so this low
edge-to-vertex ratio is not surprising.  The `data.frame` resulting from
`weight_streetnet()` is what `dodgr` uses to calculate shortest path routes, as
will be described below, following a brief description of weighting street
networks.


## Graph Components

The foregoing `graph` object returned from `weight_streetnet()` also includes a
`$component` column enumerating all of the distinct inter-connected components of
the graph.
```{r, eval = FALSE}
head (table (graph$component))
```
```{r, echo = FALSE}
cn <- c (182420, 1484, 416, 306, 148, 132)
names (cn) <- 1:6
cn
```
Components are numbered in order of decreasing size, with `$component = 1`
always denoting the largest component. In this case, that component contains
182,420 edges, representing 97.4% of the graph, with the remainder divided
between five additional components.  Component numbers can be determined for any
types of graph with the `dodgr_components()` function. For example, the
following lines reduce the previous graph to a minmial (non-spatial) structure
of four columns, and then (re-)calculate a fifth column of `$component`s:
```{r, eval = FALSE}
cols <- c ("edge_id", "from_id", "to_id", "d")
graph_min <- graph [, which (names (graph) %in% cols)]
graph_min <- dodgr_components (graph_min)
head (graph_min)
```
```{r, echo = FALSE}
from_id <- c (13060397, 13060402, 13060402, 13060405, 13060405, 13060413)
to_id <- c (13060402, 13060397, 13060405, 13060402, 13060413, 13060405)
d <- c (0.006909587, 0.006909587, 0.013732955, 0.013732955, 0.013148026,
        0.013148026)
data.frame (edge_id = 1:6,
            from_id = from_id,
            to_id = to_id,
            d = d,
            component = rep (1, 6))
```
The `component` column column can be used to select or filter any component in
a graph.  It is particularly useful to ensure routing calculations consider only
connected vertices through simply removing all minor components:
```{r}
graph_connected <- graph [graph$component == 1, ]
```
This is explored further below (under Distance Matrices).

## Weighting Profiles

Dual-weights for street networks are generally obtained by multiplying the
distance of each segment by a weighting factor reflecting the type of highway.
As demonstrated above, this can be done easily within `dodgr` with the
`weight_streetnet()` function, which using the named weighting profiles included
with the `dodgr` package. Each profile assigns weights to each distinct type of
highway.
```{r}
names (weighting_profiles)
class (weighting_profiles)
unique (weighting_profiles$name)
weighting_profiles [weighting_profiles$name == "foot", ]
```
Each profile is defined by a series of percentage weights quantifying
highway-type preferences for a particular mode of travel.  The distinct types of
highways within the York graph obtained above can be tabulated with:
```{r, eval = FALSE}
table (graph$highway)
```
```{r, echo = FALSE}
tb <- c (3122, 222, 11079, 35518, 106, 5856, 381, 2815, 23, 60, 41369, 308,
         4016, 8, 32077, 734, 15006, 26, 16452, 3144, 371, 14607)
names (tb) <- c ("bridleway", "construction", "cycleway", "footway",
                 "living_street", "path", "pedestrian", "primary",
                 "primary_link", "raceway", "residential", "road", "secondary",
                 "secondary_link", "service", "steps", "tertiary",
                 "tertiary_link", "track", "trunk", "trunk_link",
                 "unclassified")
tb
```

## Random Sub-Graphs

The `dodgr` packages includes a function to select a random connected portion of
graph including a specified number of vertices. This function is used in the
`compare_heaps()` function described below, but is also useful for general
statistical analyses of large graphs which may otherwise take too long to
compute.
```{r, eval = FALSE}
graph_sub <- dodgr_sample (graph, nverts = 1000)
nrow (graph_sub)
```
```{r, echo = FALSE}
2071
```
The random sample has around twice as many edges as vertices, in accordance with
the statistics calculated above.
```{r, eval = FALSE}
nrow (dodgr_vertices (graph_sub))
```

# Distance Matrices: `dodgr_dists()`

As demonstrated at the outset, an entire network can simply be submitted to
`dodgr_dists()`, in which case a square matrix will be returned containing
pair-wise distances between all vertices. Doing that for the `graph` of York
will return a square matrix 90,015-times-90,015 distances (around 8 billion
distances). It might be possible to do that on some computers, but is possibly
neither recommended nor desirable.

The `dodgr_dists()` function accepts additional arguments of `from` and `to`
defining points from and to which distances are to be calculated. If only `from`
is provided, a square matrix is returned of pair-wise distances between all
listed points.

## Aligning Routing Points to Graphs

For spatial graphs---that is, those containing columns of latitudes and
longitudes (or "x" and "y")---routing points can be represented by a simple
matrix of arbitrary latitudes and longitudes (or, again, "x" and "y").
`dodgr_dists()` will map these points to the closest network points, and return
corresponding shortest-path distances. This may be illustrated by repeating the
above example which generated `xy` as a two-column matrix of 1,000 coordinates,
```{r, eval = FALSE}
d <- dodgr_dists (graph, from = xy)
dim (d); range (d, na.rm = TRUE)
```
```{r, echo = FALSE}
c (1000, 1000)
c (0.00000, 35.51991)
```
If the `to` argument is also specified, the matrix returned will have rows
matching `from` and columns matching `to`
```{r, eval = FALSE}
d <- dodgr_dists (graph, from = xy, to = xy [1:10])
dim (d)
```
```{r, echo = FALSE}
c (1000, 10)
```
Some of the resultant distances in the above cases are `NA` because the points
were sampled from the entire bounding box, and the street network near the
boundaries may be cut off from the rest. As demonstrated above, the
`weight_streetnet()` function returns a `component` vector, and such
disconnected edges will have `graph$component > 1`, because `graph$component ==
1` always denotes the largest connected component.  This means that the graph
can always be reduced to the single largest component with the following single
line:
```{r, eval = FALSE}
graph_connected <- graph [graph$component == 1, ]
```
A distance matrix obtained from running `dodgr_dists` on `graph_connected` will
then contain no NA values. Note, however, that spatial routing points (expressed
as `from` and/or `to` arguments) will in this case be mapped to the nearest
vertices of `graph_connected`, rather than the potentially closer nearest points
of the full `graph`. This may make the spatial mapping of routing points less
accurate that results obtained by repeating extraction of the street network
using an expanded bounding box. For automatic extraction of street networks with
`dodgr_dists()`, the extent by which the bounding box exceeds the range of
routing points (`from` and `to` arguments) is determined by an extra parameter
`expand`, quantifying the relative extent to which the bounding box should
exceed the spatial range of the routing points.

For non-spatial graphs, `from` and `to` must match precisely on to vertices
named in the graph itself. In the graph considered above, these vertex names
were contained in the columns, `from_id` and `to_id`. The minimum that a `dodgr`
graph requires is,
```{r, eval = FALSE}
head (graph [names (graph) %in% c ("edge_id", "from_id", "to_id", "d")])
```
```{r, echo = FALSE}
edge_id <- 1:6
from_id <- c (13060397, 13060402, 13060402, 13060405, 13060405, 13060413)
to_id <- c (13060402, 13060397, 13060405, 13060402, 13060413, 13060405)
d <- c (0.006909587, 0.006909587, 0.013732955, 0.013732955, 0.013148026,
        0.013148026)
data.frame (edge_id = 1:6, from_id = as.character (from_id),
            to_id = as.character (to_id), d = d)
```
in which case `from` (and `to`, if given) must directly name the vertices in
`from_id` and `to_id`.

## Shortest Path Calculations: Priority Queues

`dodgr` uses an internal library for the calculation of shortest paths using a
variety of priority queues. These determine the order in which a graph is
traversed, and the choice of priority queue can have a considerable effect on
computational efficiency for different kinds of graphs. In contrast to `dodgr`,
most other **R** packages for shortest path calculations do not use priority
queues, and so will generally be less efficient. The difference
that priority queues can make may be demonstrated through comparison with
equivalent routines from the `igraph` package. To do so, we first need to
convert the preceding `graph` into an equivalent `igraph` structure:
```{r, eval = FALSE}
edges <- cbind (graph$from_id, graph$to_id)
nodes <- unique (as.vector (edges)) # used below in test comparison
edges <- as.vector (t (edges))
igr <- igraph::make_directed_graph (edges)
igraph::E (igr)$weight <- graph$d_weighted
```
Even then, trying to calculating routing distances with `igraph` is likely to
fail
```{r, eval = FALSE}
d <- igraph::distances (igr, v = nodes, to = nodes, mode = "out")
```
```{r, echo = FALSE}
message (paste0 ('Error in igraph::distances(igr, v = nodes, to = nodes, ',
                 'mode = "out") :\n   At vector.pmt:442 : cannot reserve ',
                 'space for vector, Out of memory'))
```
`dodgr` will perform this calculation, even though `igraph` will not. However,
an even more useful `dodgr` function is the ability to directly compare the
performance of different kinds of priority queue structures, and to compare
these `dodgr` implementations directly with `igraph` performance. This is
achieved with the `compare_heaps` function.

The comparisons are actually made on a randomly selected sub-component of the
graph containing a defined number of vertices.
```{r, eval = FALSE}
compare_heaps (graph)
```
```{r, echo = FALSE}
test <- c ('d <- dodgr_dists(graph, heap = "BHeap")',
           'd <- igraph::distances(igr, v = nodes, to = nodes, mode = "out")',
           'd <- dodgr_dists(graph, heap = "FHeap")',
           'd <- dodgr_dists(graph, heap = "TriHeap")',
           'd <- dodgr_dists(graph, heap = "TriHeapExt")',
           'd <- dodgr_dists(graph, heap = "Heap23")')
replications <- rep (10, length (test))
elapsed <- c (0.234, 0.261, 0.330, 0.360, 0.370, 0.417)
relative <- c (1.000, 1.115, 1.410, 1.538, 1.581, 1.782)
user.self <- c (0.236, 0.260, 0.328, 0.360, 0.368, 0.420)
sys.self <- user.child <- sys.child <- rep (0, length (test))
data.frame (test = test, replications = replications, elapsed = elapsed,
            relative = relative, user.self = user.self, sys.self = sys.self)
```
The key column of that `data.frame` is `relative`, which quantifies the relative
performance of each test in relation to the best which is given a score of 1.
`dodgr` using the default `heap = "BHeap"`, which is a binary heap priority
queue, performs faster than `igraph` for these graphs. Different kind of graphs
will perform differently with different priority queue structures, and this
function enables users to empirically discern the optimal heap for their kind of
graph.

Note, however, that this is not an entirely fair comparison, because `dodgr`
calculates dual-weighted distances, whereas `igraph`---and indeed all other
**R** packages---only directly calculate distances based on a single set of
weights. Implementing dual-weighted routing in these cases requires explicitly
re-tracing all paths and summing the second set of weights along each path. A
time comparison in that case would be very strongly in favour of `dodgr`.

# Graph Contraction

A further unique feature of `dodgr` is the ability to remove redundant vertices
from graphs, thereby speeding up routing calculations (see Fig. 3).

```{r, fig.align="center", echo=FALSE, engine='tikz',fig.ext='svg', code=readLines("fig3.tex"), fig.cap="Figure 3: **(A)** Vertex#2 is redundant for routing calculations from anywhere other than that point itself. (**B**) Equivalent graph to (**A**) contracted through removal of vertex#2. (**C**) Vertex#2 is not redundant."}
```

In Fig. 3(A), the only way to get from vertex 1 to 3, 4 or 5 is through C. The
intermediate vertex B is redundant for routing purposes and may simply be
removed, with directional edges inserted directly between vertices 1 and 3. This
yields the equivalent contracted graph of Fig. 3(B), in which, for example, the
distance (or weight) between 1 and 3 is the sum of previous distances (or
weights) between 1 $\to$ 2 and 2 $\to$ 3. Note that if one of the two edges
between, say, 3 and 2 were removed, vertex 2 would no longer be redundant (Fig.
3(C)).

Different kinds of graphs have different degrees of redundancy, and even street
networks differ, through for example dense inner-urban networks generally being
less redundant than less dense extra-urban or rural networks. The contracted
version of a graph can be obtained with the function `dodgr_compact_graph()`,
illustrated here with the York example from above.
```{r, eval = FALSE}
graph_compact <- dodgr_compact_graph (graph)
```
The function `dodgr_compact_graph()` returns a list with both the original and
compact graphs. Relative sizes are
```{r, eval = FALSE}
nrow (graph_compact$original); nrow (graph_compact$compact)
```
```{r, echo = FALSE}
c (187274, 112190)
```
equivalent to the removal of just over 40% of all edges. The difference in
routing efficiency can be seen with the following code
```{r, eval = FALSE}
rbenchmark::benchmark (
                       d1 <- dodgr_dists (graph_compact$compact, from = xy),
                       d2 <- dodgr_dists (graph, from = xy),
                       replications = 2)
```
```{r, echo = FALSE}
test <- c ('d1 <- dodgr_dists(graph_compact$compact, from = xy)',
           'd2 <- dodgr_dists(graph, from = xy)')
replications <- c (2, 2)
elapsed <- c (3.574, 47.824)
relative <- c (1.000, 13.381)
user.self <- c (3.576, 47.769)
sys.self <- user.child <- sys.child <- rep (0, length (test))
data.frame (test = test, replications = replications, elapsed = elapsed,
            relative = relative, user.self = user.self, sys.self = sys.self)
```
And reducing numbers of graph edges by only 40% enables pairwise routing between
1,000 points through 1-200,000 vertices to be conducted 13 times faster. All routing
algorithms scale non-linearly with size, and relative improvements in efficiency
will be even greater for larger graphs. Note that these graphs of just a couple
of hundred thousand edges will generally not be able to be processed in
`igraph`, yet are relatively small in `dodgr` terms. Although maximal graph
sizes capable of being processed in `dodgr` will of course be machine dependent,
typical values will be at least a few million edges, and possibly much larger.

