---
title: "dodgr isochrones, isodists, and isoverts"
author: "Mark Padgham"
date: "`r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: false
        theme: flatly
vignette: >
  %\VignetteIndexEntry{4 dodgr-iso}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, echo = FALSE}
opts <- knitr::opts_chunk$get ()
knitr::opts_chunk$set (fig.height = 6)
```

```{r pkg-load, echo = FALSE, message = FALSE}
library (dodgr)
```

The `dodgr`package includes three functions for calculating iso-contours from
one or more points of origin:

- [`dodgr_isodists()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isodists.html)
for contours of equal distance;
- [`dodgr_isochrones()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isochrones.html)
for contours of equal time; and
- [`dodgr_isoverts()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isoverts.html)
to return full set of vertices within defined isodistance or isochrone contours.

Each of these functions is fully vectorised to accept both multiple origin
points and multiple contours. Each returns a single `data.frame` object, with
a column specifying either the distance or time limit (as `dlim` for 
[`dodgr_isodists()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isodists.html),
or `tlim` for
[`dodgr_isochrones()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isochrones.html)).
The functions are also internally parallelised for efficient calculation.

## Input formats

The [`dodgr_isodists()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isodists.html)
function accepts arbitrary input, either as a generic `dodgr` `data.frame`
(with a "distance" column), or as street networks derived from applying the
[`weight_streetnet()`](https://UrbanAnalyst.github.io/dodgr/reference/weight_streetnet.html)
function to either [`sf`](https://r-spatial.github.io/sf) or
[`st`](https://hypertidy.github.io/silicate) objects. Time-based calculations
in `dodgr` are only possible when applied to street networks dervied from 
[`st`](https://hypertidy.github.io/silicate) objects, because equivalent
`sf`-based objects can only provide crude estimates of journey times. The time-specific functions 
[`dodgr_isochrones()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isochrones.html)
and
[`dodgr_isoverts()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isoverts.html)
thus only accept `dodgr` street networks derived from `sc` data. The following
table summarises the requirements of these functions:

Function | Input type
--- | ---
dodgr_isodists | Either `sf` or `sc` format data
dodgr_ isochrones | Only `sc` format data
dodgr_isoverts | Only `sc` format data


## dodgr_isodists

The
[`dodgr_isodists()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isodists.html)
function calculates contours of equal distance from one or more points of origin.
```{r isodists}
graph <- weight_streetnet (hampi)
from <- sample (graph$from_id, size = 100)
dlim <- c (2, 5, 10, 20) * 100
d <- dodgr_isodists (graph, from = from, dlim)
dim (d)
knitr::kable (head (d))
```

This function returns in this case a `data.frame` with the columns shown above
and `r format (nrow (d), big.mark = ",")` points defining the isodistance
contours from each origin (`from`) point, at the specified distances of
`r paste0 (dlim, collapse = ", ")` metres. There are naturally more points
defining the contours at longer distances:
```{r isodist-table}
table (d$dlim)
```
The points defining the contours are arranged in an anticlockwise order around
each point of origin, and so can be directly visualised using base graphics
functions. This is easier to see by using just a single point of origin:

```{r isodist-plot, echo = TRUE}
set.seed (1)
from <- sample (graph$from_id, size = 1)
dlim <- c (2, 5, 10, 20) * 100
d <- dodgr_isodists (graph, from = from, dlim)
cols <- terrain.colors (length (dlim) + 2) # + 2 to avoid light colours at end of scale.
index <- which (d$dlim == max (d$dlim)) # plot max contour first
plot (
    d$x [index], d$y [index], "l",
    col = cols [1],
    xlab = "longitude", ylab = "latitude"
)
for (i in seq (dlim) [-1]) {
    index <- which (d$dlim == rev (dlim) [i])
    lines (d$x [index], d$y [index], col = cols [i], lwd = i + 1)
}
```

### The `concavity` parameter

The default contours are convex hulls. The
[`dodgr_isodists()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isodists.html)
and
[`dodgr_isochrones()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isochrones.html)
functions include an additional `concavity` parameter to return generate concave 
isodistance or isochrone polygons, defaulting to a value of 0 for no
concavity. Changing this value can generate concave contours like in the
following results:

```{r isodist-plot-concave, echo = TRUE}
set.seed (1)
from <- sample (graph$from_id, size = 1)
dlim <- c (2, 5, 10, 20) * 100
d <- dodgr_isodists (graph, from = from, dlim, concavity = 0.5)
cols <- terrain.colors (length (dlim))
index <- which (d$dlim == max (d$dlim))
plot (
    d$x [index], d$y [index], "l",
    col = cols [1],
    xlab = "longitude", ylab = "latitude"
)
for (i in seq (dlim) [-1]) {
    index <- which (d$dlim == rev (dlim) [i])
    lines (d$x [index], d$y [index], col = cols [i], lwd = i + 1)
}
```

The calculation of `isodists` or `isochrone` values is cached, so different
values of the `concavity` parameter can be trialled without needing to
re-calculate the underlying values:

```{r iso-caching}
set.seed (2)
from <- sample (graph$from_id, size = 1)
dlim <- c (2, 5, 10, 20) * 100
system.time ( # Initial call calculates distances to all points:
    d <- dodgr_isodists (graph, from = from, dlim, concavity = 0.5)
)
system.time ( # Subsequent call uses cached values:
    d <- dodgr_isodists (graph, from = from, dlim, concavity = 0.75)
)
```

The `concavity` parameter has a default value of 0 for strictly convex
polygons. Values closer to 1 generate more concave polygons. The following code
demonstrates the effect of increasing values of this parameter.

```{r concavity}
set.seed (5)
from <- sample (graph$from_id, size = 1)
dlim <- 2000
concavity <- 0:4 / 4
d <- lapply (concavity, function (i) dodgr_isodists (graph, from = from, dlim, concavity = i))
cols <- terrain.colors (length (d) + 2)
plot (
    d [[1]]$x, d [[1]]$y,
    type = "l",
    col = cols [1], lwd = 5,
    xlab = "longitude", ylab = "latitude"
)
for (i in seq_along (d) [-1]) {
    lines (
        d [[i]]$x, d [[i]]$y,
        col = cols [i], lwd = 6 - i, lty = i
    )
}
v <- dodgr_vertices (graph)
v_from <- v [match (from, v$id), ]
points (v_from [, c ("x", "y")], pch = 19, col = "red", cex = 2)
```

```{r knitr-opts-reset, echo = FALSE}
opts <- knitr::opts_chunk$get ()
knitr::opts_chunk$set (fig.height = opts$fig.height)
```

## dodgr_isochrones

The
[`dodgr_isochrones()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isochrones.html)
works just like
[`dodgr_isodists()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isodists.html)
except that it can only be applied to street networks generated from
[`silicate` (SC) format](https://hypertidy.github.io/silicate) data.
The results differ only from equivalent distance results in having a column
named "tlim" (for "time") instead of "dlim" (for "distance"). The `concavity`
parameter functions in exactly the same way as described above.

## dodgr_isoverts

Both the
[`dodgr_isodists()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isodists.html)
and
[`dodgr_isochrones()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isochrones.html)
functions work by estimating distances or times from specified origin points to
all points out to the maximum specified values of `dlim` or `tlim`. A second
stage then generates convex or concave hulls around these points, with the
result containing an ordered sequence of points tracing the contours around
these hulls.

The
[`dodgr_isoverts()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isoverts.html)
function generates the full underlying set of points without applying any
hull-tracing routines. This resultant points may be used, for example, to apply
some custom hull-tracing routine, or to perform statistical analyses on the
full points enclosed within specified values of `dlim` or `tlim`. The object
returned from this function is identical to those demonstrated above, with
a column named "dlim" if the `dlim` parameter is specified, or "tlim" if the
`tlim` parameter is specified. The `data.frame` returned from
[`dodgr_isoverts()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isoverts.html)
will always contain more data, and have more rows, than equivalent results from
[`dodgr_isodists()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isodists.html)
or
[`dodgr_isochrones()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isochrones.html).
The rows of the `data.frame` returned from
[`dodgr_isoverts()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isoverts.html)
are ordered by increasing `dlim` or `tlim` values, but rows for any given value
of those parameters are in arbitrary order. Each successively higher value of either
`dlim` or `tlim` includes all points denoted by all lower values. For example,
all values contained within `dlim = 200` are all those with that specified
value of `dlim` *plus* all rows with any values of `dlim < 200`.

The following code demonstrates typical sizes of results from the
[`dodgr_isodists()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isodists.html)
and
[`dodgr_isoverts()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isoverts.html)
functions:

```{r nrow-comparison1, message = FALSE, echo = TRUE, eval = FALSE}
dat <- dodgr_streetnet_sc ("hampi india")
graph <- weight_streetnet (dat, wt_profile = "foot")
set.seed (1)
from <- sample (graph$.vx0, size = 1)
dlim <- c (2, 5, 10, 20) * 100
nrow (dodgr_isodists (graph, from = from, dlim))
```
```{r nrow-output1, echo = FALSE}
34
```

```{r nrow-comparison2, echo = TRUE, eval = FALSE}
nrow (dodgr_isoverts (graph, from = from, dlim))
```
```{r nrow-output2, echo = FALSE}
145
```
