---
title: "dodgr isochrones, isodists, and isoverts"
author: "Mark Padgham"
date: "`r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: false
        theme: flatly
vignette: >
  %\VignetteIndexEntry{4 dodgr-iso}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r pkg-load, echo = FALSE, message = FALSE}
library (dodgr)
```

The `dodgr`package includes three functions for calculating iso-contours from
one or more points of origin:
[`dodgr_isodists()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isodists.html)
for contours of equal distance;
[`dodgr_isochrones()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isochrones.html)
for contours of equal time; and
[`dodgr_isoverts()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isoverts.html)
to return full set of vertices within defined isodistance or isochrone contours.

Each of these functions is fully vectorised to accept both multiple origin
points and multiple contours. Each returns a single `data.frame` object, with
a column specifying either the distance or time limit (as `dlim` for 
[`dodgr_isodists()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isodists.html),
or `tlim` for
[`dodgr_isochrones()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isochrones.html)).
The functions are also internally parallelised for efficient calculation.

## dodgr_isodists

The
[`dodgr_isodists()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isodists.html)
function calculates contours of equal distance from one or more points of origin.
```{r isodists}
graph <- weight_streetnet (hampi)
from <- sample (graph$from_id, size = 100)
dlim <- c (1, 2, 5, 10, 20) * 100
d <- dodgr_isodists (graph, from = from, dlim)
dim (d)
knitr::kable (head (d))
```

This function returns in this case a `data.frame` with the columns shown above
and `r format (nrow (d), big.mark = ",")` points defining the isodistance
contours from each origin (`from`) point, at the specified distances of
`r paste0 (dlim, collapse = ", ")` metres. There are naturally more points
defining the contours at longer distances:
```{r isodist-table}
table (d$dlim)
```
The points defining the contours are arranged in an anticlockwise order around
each point of origin, and so can be directly visualised using base graphics
functions. This is easier to see by using just a single point of origin:

```{r isodist-plot-fakey, eval = FALSE}
set.seed (42)
from <- sample (graph$from_id, size = 1)
dlim <- c (1, 2, 5, 10, 20) * 100
d <- dodgr_isodists (graph, from = from, dlim)
cols <- terrain.colors (length (dlim))
index <- which (d$dlim == max (d$dlim)) # plot max contour first
plot (
    d$x [index], d$y [index], "l",
    col = cols [1],
    xlab = "longitude", ylab = "latitude"
)
for (i in seq (dlim) [-1]) {
    index <- which (d$dlim == rev (dlim) [i])
    lines (d$x [index], d$y [index], col = cols [i], lwd = i + 1)
}
```
```{r isodist-plot, echo = FALSE}
set.seed (1)
from <- sample (graph$from_id, size = 1)
dlim <- c (1, 2, 5, 10, 20) * 100
d <- dodgr_isodists (graph, from = from, dlim)
cols <- terrain.colors (length (dlim))
index <- which (d$dlim == max (d$dlim)) # plot max contour first
plot (
    d$x [index], d$y [index], "l",
    col = cols [1],
    xlab = "longitude", ylab = "latitude"
)
for (i in seq (dlim) [-1]) {
    index <- which (d$dlim == rev (dlim) [i])
    lines (d$x [index], d$y [index], col = cols [i], lwd = i + 1)
}
```

The default contours are convex hulls. The
[`dodgr_isodists()`](https://UrbanAnalyst.github.io/dodgr/reference/dodgr_isodists.html)
function includes an additional `concavity` parameter to return generate concave 
isodistance (or isochrone) polygons, defaulting to a value of 0 for no
concavity. Changing this value can generate concave contours like in the
following results:

```{r isodist-plot-convex, echo = FALSE}
set.seed (1)
from <- sample (graph$from_id, size = 1)
dlim <- c (1, 2, 5, 10, 20) * 100
d <- dodgr_isodists (graph, from = from, dlim, concavity = 0.5)
cols <- terrain.colors (length (dlim))
index <- which (d$dlim == max (d$dlim)) # plot max contour first
plot (
    d$x [index], d$y [index], "l",
    col = cols [1],
    xlab = "longitude", ylab = "latitude"
)
for (i in seq (dlim) [-1]) {
    index <- which (d$dlim == rev (dlim) [i])
    lines (d$x [index], d$y [index], col = cols [i], lwd = i + 1)
}
```
